# Energy Data Visualization Project

**Author**: Marcello Russo **Date**: `r Sys.Date()`

------------------------------------------------------------------------

## Setup

Load libraries and dataset.

```{r setup}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidytuesdayR)
library(scales)
library(tidyr)
library(forcats)
library(ggrepel)

# Load data
tuesdata <- tidytuesdayR::tt_load(2023, week = 23)

energy_data <- tuesdata$'owid-energy'
energy_data <- energy_data %>%
  filter(
    !grepl("\\(", country),         # esclude "Africa (BP)", ecc.
    !grepl("-", country),           # esclude "IEO - Africa (EIA)", ecc.
    !country %in% c("Africa", "Asia", "Europe", "North America", "South America", "Oceania", "World"),
    !grepl("G7|G20|OECD|OPEC|income|region|Union|BP|EIA|Ember", country)
  )

get_continent <- function(country) {
  europe <- c("Italy", "France", "Germany", "Spain", "Portugal", "Netherlands", "Belgium",
              "Sweden", "Norway", "Finland", "Denmark", "Austria", "Switzerland", "Greece",
              "Ireland", "Poland", "Czechia", "Hungary", "Slovakia", "Slovenia",
              "Croatia", "Romania", "Bulgaria", "Estonia", "Latvia", "Lithuania", "Malta",
              "Cyprus", "Luxembourg", "Iceland", "Albania", "Belarus", "Bosnia and Herzegovina",
              "North Macedonia", "Moldova", "Montenegro", "Serbia", "Ukraine", "United Kingdom")

  asia <- c("China", "Japan", "South Korea", "India", "Indonesia", "Malaysia", "Thailand",
            "Vietnam", "Philippines", "Bangladesh", "Pakistan", "Sri Lanka", "Nepal", "Iran",
            "Iraq", "Israel", "Kazakhstan", "Kuwait", "Kyrgyzstan", "Laos", "Lebanon", "Mongolia",
            "Myanmar", "North Korea", "Oman", "Palestine", "Saudi Arabia", "Singapore", "Syria",
            "Taiwan", "Tajikistan", "Timor", "Turkey", "Turkmenistan", "United Arab Emirates",
            "Uzbekistan", "Yemen", "Armenia", "Georgia", "Azerbaijan", "Jordan", "Qatar", "Bahrain",
            "Afghanistan", "Bhutan", "Maldives", "Thailand")

  africa <- c("South Africa", "Nigeria", "Egypt", "Kenya", "Morocco", "Ghana", "Algeria",
              "Tunisia", "Ethiopia", "Ivory Coast", "Senegal", "Angola", "Botswana", "Cameroon",
              "Congo", "Democratic Republic of Congo", "Gabon", "Gambia", "Guinea", "Guinea-Bissau",
              "Madagascar", "Malawi", "Mali", "Mauritania", "Mauritius", "Mozambique", "Namibia",
              "Niger", "Rwanda", "Seychelles", "Sierra Leone", "Somalia", "Sudan", "Tanzania",
              "Togo", "Uganda", "Zambia", "Zimbabwe", "Benin", "Burkina Faso", "Burundi", "Chad",
              "Comoros", "Djibouti", "Eritrea", "Equatorial Guinea", "Eswatini", "Lesotho",
              "Liberia", "Libya", "Sao Tome and Principe", "South Sudan", "Western Sahara")

  north_america <- c("United States", "Canada", "Mexico", "Bahamas", "Barbados", "Belize",
                     "Costa Rica", "Cuba", "Dominica", "Dominican Republic", "El Salvador",
                     "Grenada", "Guatemala", "Haiti", "Honduras", "Jamaica", "Nicaragua",
                     "Panama", "Saint Kitts and Nevis", "Saint Lucia", "Saint Vincent and the Grenadines",
                     "Trinidad and Tobago")

  south_america <- c("Brazil", "Argentina", "Chile", "Colombia", "Peru", "Uruguay", "Ecuador",
                     "Paraguay", "Venezuela", "Bolivia", "Guyana", "Suriname")

  oceania <- c("Australia", "New Zealand", "Fiji", "Kiribati", "Marshall Islands", "Micronesia",
               "Nauru", "Palau", "Papua New Guinea", "Samoa", "Solomon Islands", "Tonga",
               "Tuvalu", "Vanuatu")

  if (country %in% europe) {
    return("Europe")
  } else if (country %in% asia) {
    return("Asia")
  } else if (country %in% africa) {
    return("Africa")
  } else if (country %in% north_america) {
    return("North America")
  } else if (country %in% south_america) {
    return("South America")
  } else if (country %in% oceania) {
    return("Oceania")
  } else {
    return("Unknown")
  }
}
```

------------------------------------------------------------------------

## Plot 1: GDP vs. Per Capita Energy Consumption

**Question**: Is there a relationship between GDP and per capita energy consumption?

```{r plot1}
# Data processing corretto
plot1data <- energy_data %>%
  filter(year == 2018) %>%
  mutate(continent = sapply(country, get_continent)) %>%
  filter(continent != "Unknown") %>%
  mutate(
    gdp_per_capita = gdp / population
  ) %>%
  filter(
    !is.na(gdp), 
    !is.na(population),
    !is.na(energy_per_capita),
    !is.na(continent)
  )

# Plot con scala lineare
ggplot(plot1data, aes(x = gdp_per_capita, y = energy_per_capita, color = continent)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_x_log10(
    name = "GDP per capita (USD, log scale)",
    labels = scales::dollar_format(accuracy = 1),
    breaks = c(200, 500, 1000, 2000, 5000, 10000, 20000, 50000, 100000),
    limits = c(200, 150000)
  ) +
  scale_y_log10(
    name = "Energy use per capita (kWh, log scale)",
    labels = scales::comma_format(),
    breaks = c(100, 300, 1000, 3000, 10000, 30000, 100000, 200000),
    limits = c(100, 200000)
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Energy Consumption vs Economic Output (2018)",
    subtitle = "Log-log scale clarifies the power-law distribution",
    color = "Continent"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot",
    plot.margin = margin(1, 1, 1, 1, "cm"),
    axis.title = element_text(face = "bold")
  )

ggsave("plots/energy_gdp_plot.png", width = 10, height = 7, dpi = 300, bg = "white")
```

**Explanation**:\
- We use a **scatter plot** to explore the relationship between GDP and energy consumption.\
- Points are colored by `fossil_share_energy` to show how fossil fuel dependency influences the trend.\
- Outliers (top 10% energy consumers) are labeled using `ggrepel` for clarity.

------------------------------------------------------------------------

## Plot 2: Solar and Wind Energy Leaders

**Question**: Which countries lead in solar and wind energy adoption?

```{r plot2}
# Get latest year's data for top 10 solar/wind countries
plot2_data <- energy_data %>%
  filter(year == latest_year) %>%
  select(country, solar_share_elec, wind_share_elec) %>%
  pivot_longer(cols = -country, names_to = "source", values_to = "share") %>%
  group_by(source) %>%
  slice_max(order_by = share, n = 10) 

# Horizontal dodged bar chart
ggplot(plot2_data, aes(x = share, y = country, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("solar_share_elec" = "#FFD700", "wind_share_elec" = "#4682B4"),
    labels = c("Solar", "Wind")
  ) +
  labs(
    title = "Top 10 Countries in Solar and Wind Energy Adoption",
    x = "Share of Electricity Generation (%)", 
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Explanation**:\
- A **dodged bar chart** compares the top 10 countries for solar and wind energy.\
- Data is pivoted to long format for easy faceting.\
- Gold (`#FFD700`) represents solar, blue (`#4682B4`) represents wind.

------------------------------------------------------------------------

## Plot 3: Population Growth vs. Energy Demand

**Question**: Do countries with higher population growth experience larger changes in energy demand?

```{r plot3}
# Calculate percentage changes over 10 years
plot3_data <- energy_data %>%
  group_by(country) %>%
  filter(year >= latest_year - 10) %>%
  summarise(
    pop_change = (last(population) - first(population)) / first(population) * 100,
    demand_change = (last(electricity_demand) - first(electricity_demand)) / first(electricity_demand) * 100
  ) %>%
  drop_na()

# Scatter plot with regression
ggplot(plot3_data, aes(x = pop_change, y = demand_change)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Population Growth vs. Electricity Demand Change",
    x = "Population Change (%) (10-year delta)",
    y = "Electricity Demand Change (%)"
  ) +
  theme_minimal()
```

**Explanation**:\
- Percentage changes in population and electricity demand are calculated over a 10-year window.\
- A **regression line** (`geom_smooth`) shows the overall trend.\
- Points represent individual countries.

------------------------------------------------------------------------

## Plot 4: Energy Consumption vs. Efficiency

**Question**: Do countries with high per capita energy consumption also have high energy efficiency?

```{r plot4}
# Calculate medians for quadrants
median_energy <- median(plot1_data$energy_per_capita, na.rm = TRUE)
median_efficiency <- median(plot1_data$energy_per_gdp, na.rm = TRUE)

ggplot(plot1_data, aes(x = energy_per_capita, y = energy_per_gdp, color = continent)) +
  geom_point() +
  geom_hline(yintercept = median_efficiency, linetype = "dashed") +
  geom_vline(xintercept = median_energy, linetype = "dashed") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Energy Consumption vs. Efficiency",
    x = "Energy per Capita (kWh)",
    y = "Energy per GDP (kWh/$)"
  ) +
  theme_minimal()
```

**Explanation**:\
- **Quadrant analysis** splits the plot into four regions using median values.\
- Points are colored by continent to highlight regional patterns.\
- High consumption + low efficiency (top-left quadrant) may indicate wasteful economies.

------------------------------------------------------------------------

## Plot 5: Carbon Intensity vs. Fossil Dependency

**Question**: Which countries have high carbon intensity despite low fossil fuel dependency?

```{r plot5}
plot5_data <- energy_data %>%
  filter(year == latest_year) %>%
  drop_na(carbon_intensity_elec, fossil_share_energy)

ggplot(plot5_data, aes(x = fossil_share_energy, y = carbon_intensity_elec, color = continent)) +
  geom_point() +
  geom_text_repel(
    data = subset(plot5_data, carbon_intensity_elec > 500 & fossil_share_energy < 50),
    aes(label = country), size = 3
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Carbon Intensity vs. Fossil Fuel Dependency",
    x = "Fossil Fuel Share of Energy (%)",
    y = "Carbon Intensity (gCO₂/kWh)"
  ) +
  theme_minimal()
```

**Explanation**:\
- Outliers (e.g., countries with `carbon_intensity_elec > 500` but `fossil_share_energy < 50`) are labeled.\
- These countries may rely on coal despite having "low" fossil fuel shares.

------------------------------------------------------------------------

## Plot 6: Energy Mix by Continent

**Question**: How does the energy mix vary by continent?

```{r plot6}
plot6_data <- energy_data %>%
  filter(year == latest_year) %>%
  group_by(continent) %>%
  summarise(
    fossil = mean(fossil_share_energy, na.rm = TRUE),
    renewables = mean(renewables_share_energy, na.rm = TRUE),
    nuclear = mean(nuclear_share_energy, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -continent, names_to = "source", values_to = "share")

ggplot(plot6_data, aes(x = continent, y = share, fill = source)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("fossil" = "#E69F00", "renewables" = "#009E73", "nuclear" = "#0072B2")) +
  labs(
    title = "Energy Mix by Continent",
    x = "",
    y = "Proportion of Energy Mix"
  ) +
  theme_minimal()
```

**Explanation**:\
- A **stacked bar chart** shows the proportional contribution of each energy source.\
- Fossil fuels dominate in some continents, while others have more balanced mixes.

------------------------------------------------------------------------

## Saving the Notebook

Render the notebook to HTML/PDF by clicking "Knit" in RStudio.
