# Energy Data Visualization Project

**Author**: Marcello Russo **Date**: `r Sys.Date()`

------------------------------------------------------------------------

## Setup

Load libraries and dataset.

```{r setup}
# Load required libraries
library(ggplot2)
library(dplyr)
library(ggrepel) # For non-overlapping labels
library(countrycode) # To assign continents
library(tidyr)

# Load data
tuesdata <- tidytuesdayR::tt_load(2023, week = 23)

energy_data <- tuesdata$`owid-energy`

# Add continent column (if not already present)
energy_data <- energy_data %>%
  mutate(continent = countrycode(country, "country.name", "continent"))
```

------------------------------------------------------------------------

## Plot 1: GDP vs. Per Capita Energy Consumption

**Question**: Is there a relationship between GDP and per capita energy consumption?

```{r plot1}
# Filter latest year and remove NA
latest_year <- max(energy_data$year, na.rm = TRUE)
plot1_data <- energy_data %>%
  filter(year == latest_year) %>%
  drop_na(gdp, energy_per_capita, fossil_share_energy)

# Scatter plot with enhancements
ggplot(plot1_data, aes(x = gdp, y = energy_per_capita, color = fossil_share_energy)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    data = subset(plot1_data, energy_per_capita > quantile(energy_per_capita, 0.9)),
    aes(label = country), size = 3
  ) +
  scale_color_viridis_c(name = "Fossil Fuel Share (%)") +
  labs(
    title = "GDP vs. Energy Consumption per Capita",
    x = "GDP (inflation-adjusted)", 
    y = "Energy Consumption per Capita (kWh)"
  ) +
  theme_minimal()
```

**Explanation**:\
- We use a **scatter plot** to explore the relationship between GDP and energy consumption.\
- Points are colored by `fossil_share_energy` to show how fossil fuel dependency influences the trend.\
- Outliers (top 10% energy consumers) are labeled using `ggrepel` for clarity.

------------------------------------------------------------------------

## Plot 2: Solar and Wind Energy Leaders

**Question**: Which countries lead in solar and wind energy adoption?

```{r plot2}
# Get latest year's data for top 10 solar/wind countries
plot2_data <- energy_data %>%
  filter(year == latest_year) %>%
  select(country, solar_share_elec, wind_share_elec) %>%
  pivot_longer(cols = -country, names_to = "source", values_to = "share") %>%
  group_by(source) %>%
  slice_max(order_by = share, n = 10) 

# Horizontal dodged bar chart
ggplot(plot2_data, aes(x = share, y = country, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("solar_share_elec" = "#FFD700", "wind_share_elec" = "#4682B4"),
    labels = c("Solar", "Wind")
  ) +
  labs(
    title = "Top 10 Countries in Solar and Wind Energy Adoption",
    x = "Share of Electricity Generation (%)", 
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Explanation**:\
- A **dodged bar chart** compares the top 10 countries for solar and wind energy.\
- Data is pivoted to long format for easy faceting.\
- Gold (`#FFD700`) represents solar, blue (`#4682B4`) represents wind.

------------------------------------------------------------------------

## Plot 3: Population Growth vs. Energy Demand

**Question**: Do countries with higher population growth experience larger changes in energy demand?

```{r plot3}
# Calculate percentage changes over 10 years
plot3_data <- energy_data %>%
  group_by(country) %>%
  filter(year >= latest_year - 10) %>%
  summarise(
    pop_change = (last(population) - first(population)) / first(population) * 100,
    demand_change = (last(electricity_demand) - first(electricity_demand)) / first(electricity_demand) * 100
  ) %>%
  drop_na()

# Scatter plot with regression
ggplot(plot3_data, aes(x = pop_change, y = demand_change)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Population Growth vs. Electricity Demand Change",
    x = "Population Change (%) (10-year delta)",
    y = "Electricity Demand Change (%)"
  ) +
  theme_minimal()
```

**Explanation**:\
- Percentage changes in population and electricity demand are calculated over a 10-year window.\
- A **regression line** (`geom_smooth`) shows the overall trend.\
- Points represent individual countries.

------------------------------------------------------------------------

## Plot 4: Energy Consumption vs. Efficiency

**Question**: Do countries with high per capita energy consumption also have high energy efficiency?

```{r plot4}
# Calculate medians for quadrants
median_energy <- median(plot1_data$energy_per_capita, na.rm = TRUE)
median_efficiency <- median(plot1_data$energy_per_gdp, na.rm = TRUE)

ggplot(plot1_data, aes(x = energy_per_capita, y = energy_per_gdp, color = continent)) +
  geom_point() +
  geom_hline(yintercept = median_efficiency, linetype = "dashed") +
  geom_vline(xintercept = median_energy, linetype = "dashed") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Energy Consumption vs. Efficiency",
    x = "Energy per Capita (kWh)",
    y = "Energy per GDP (kWh/$)"
  ) +
  theme_minimal()
```

**Explanation**:\
- **Quadrant analysis** splits the plot into four regions using median values.\
- Points are colored by continent to highlight regional patterns.\
- High consumption + low efficiency (top-left quadrant) may indicate wasteful economies.

------------------------------------------------------------------------

## Plot 5: Carbon Intensity vs. Fossil Dependency

**Question**: Which countries have high carbon intensity despite low fossil fuel dependency?

```{r plot5}
plot5_data <- energy_data %>%
  filter(year == latest_year) %>%
  drop_na(carbon_intensity_elec, fossil_share_energy)

ggplot(plot5_data, aes(x = fossil_share_energy, y = carbon_intensity_elec, color = continent)) +
  geom_point() +
  geom_text_repel(
    data = subset(plot5_data, carbon_intensity_elec > 500 & fossil_share_energy < 50),
    aes(label = country), size = 3
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Carbon Intensity vs. Fossil Fuel Dependency",
    x = "Fossil Fuel Share of Energy (%)",
    y = "Carbon Intensity (gCO₂/kWh)"
  ) +
  theme_minimal()
```

**Explanation**:\
- Outliers (e.g., countries with `carbon_intensity_elec > 500` but `fossil_share_energy < 50`) are labeled.\
- These countries may rely on coal despite having "low" fossil fuel shares.

------------------------------------------------------------------------

## Plot 6: Energy Mix by Continent

**Question**: How does the energy mix vary by continent?

```{r plot6}
plot6_data <- energy_data %>%
  filter(year == latest_year) %>%
  group_by(continent) %>%
  summarise(
    fossil = mean(fossil_share_energy, na.rm = TRUE),
    renewables = mean(renewables_share_energy, na.rm = TRUE),
    nuclear = mean(nuclear_share_energy, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -continent, names_to = "source", values_to = "share")

ggplot(plot6_data, aes(x = continent, y = share, fill = source)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("fossil" = "#E69F00", "renewables" = "#009E73", "nuclear" = "#0072B2")) +
  labs(
    title = "Energy Mix by Continent",
    x = "",
    y = "Proportion of Energy Mix"
  ) +
  theme_minimal()
```

**Explanation**:\
- A **stacked bar chart** shows the proportional contribution of each energy source.\
- Fossil fuels dominate in some continents, while others have more balanced mixes.

------------------------------------------------------------------------

## Saving the Notebook

Render the notebook to HTML/PDF by clicking "Knit" in RStudio.
